---
title: "Trees"
author: "Gökce Ergün"
date: "2/3/2021"
output: html_document

---
Hello world

Libraries that are used
```{r, warning=FALSE}
#install.packages('dplyr')
library(dplyr)
#install.packages('lubridate')
library(lubridate)
#install.packages('tidyr')
library(tidyr)
#install.packages('ggplot2')
library(ggplot2)
library(tidyverse)
#install.packages('gganimate')
library(gganimate)
#install.packages('gifski')
library(gifski)
#install.packages('av')
library(av)
#install.packages('ggmap')
library(ggmap)
#install.packages('shiny')
library(shiny)
#install.packages('leaflet')
library(leaflet)
```



# Reading the data
```{r}
df_trees_orig <- read.csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-28/sf_trees.csv')

```

We create a copy of data to preserve its original form in case we want to look at it for reference
```{r}
df_trees <- df_trees_orig
```

# Data Cleaning 

We look at the data structure, and see all the string variables are saved as char. We would rather have them as factors for potential modelling use.
```{r}
str(df_trees)
```

Changing columns with type character to type factor. you need double brackets because single bracket indexing returns a df object,
which always evaluates to FALSE. Double brackets returns you the column as a vector whose data type can be tested. Alternatively, we can use single brackets but explicitly calling all rows and columns i.e [,i]
```{r}

for (i in 1:12){ 
    if (is.character(df_trees[,i]) == TRUE) {
    df_trees[,i]<-as.factor(df_trees[,i])
  }
  }

```

<!--check the changes worked-->

```{r, include=FALSE}
str(df_trees)
``` 

date column has char data type. We change it to date. 
```{r}
# df_trees$date <- as.Date(df_trees$date,format="%Y-%m-%d") #using base r

df_trees$date <- lubridate::ymd(df_trees$date) #using lubridate
```

save the planted year of the tree as a separate column
```{r}
# df_trees$year <- format(df_trees$date, "%Y")  #using base r

df_trees$year <- year(df_trees$date) # using lubridate
class(df_trees$year)
```

take a look at the format of species name in species column. 
```{r}
head(unique(df_trees$species))

```
we see that for each **species**, the species column reports both the latin and common names. For future analysis, it might be more 
convenient to split the latin and common names into respective columns. 

Separate the species column into latin and common name columns.
```{r}
df_trees <- df_trees %>% 
  separate(species, sep = "::", remove = FALSE, into = c('species_lat', 'species_nor'))
```

<!--check the name split worked-->
```{r, include=FALSE}
head(df_trees[,c('species','species_lat','species_nor')])
```

Check the percentage of missing values in each column

```{r}
for (col in colnames(df_trees)){
  print(paste(col, "=", mean(is.na(df_trees[[col]])))) #mean takes percentage here because adds up all TRUES (since their value is 1)
                                                       #and divides by total number of values (TRUE + FALSE)
}
```
Column with most significant amount of missing data is the **date** column.

We see almost every species has missing values (517 out of 571 species) so there is at least no clear indication that the data is missing not at random.
```{r}
df_trees %>% filter(is.na(date))%>%
  select(species)%>%
  unique() %>% dim()
 
```

further analysis of missing date values: percentage of observations with missing date values, per species.
```{r}
df_trees_datena <- df_trees %>% filter(is.na(date)) %>% 
  select(c(species, date)) %>%
  count(species)
  

df_trees_species_count <- df_trees %>% count(species)

df_trees_merge <- dplyr::left_join(df_trees_species_count, df_trees_datena, by="species", suffix = c("total", "date.na"))
df_trees_merge$percent.missing <- df_trees_merge$ndate.na / df_trees_merge$ntotal
df_trees_merge %>%
  arrange(desc(ntotal), desc(percent.missing))

```



We look at the total number of observations we would lose if we drop species where missing 100% of date information. And the number is very small relative to total number of observations (n = 192987) 

```{r}
df_trees_merge %>% filter(percent.missing==1) %>% select(ntotal) %>%
  sum()
  #arrange(desc(percent.missing), desc(ntotal))
```

How the percentage of missing date values per species is distributed. All species that have 100% of date data missing have a very low number of observations (n <=100).    

```{r}
df_trees_merge %>% 
  ggplot(aes(percent.missing))+geom_histogram() +scale_x_continuous(breaks=seq(0,1, .10))

```


Therefore, we now filter the cases based on this 100 ntotal threshold to get a cleaner histogram. 

```{r}
df_trees_merge %>% filter(ntotal>100) %>%
  ggplot(aes(percent.missing))+geom_histogram() +scale_x_continuous(breaks=seq(0,1, .10))

```


# Exploratory Analysis

Checking year of plantation range. most of the trees (50%) were planted between 2001 and 2020. However, it is likely that many of
the older trees simply were not recorded so they either dont appear in our data or have a missing date value. 
```{r}
df_trees %>%
  select(year)%>%
  quantile(na.rm = TRUE)
  
```

Create a new column that shows age using the year column to calculate it. Use the more modern approach mutate(). Then use age column 
to get an overview of the distribution of tree ages via a histogram.
```{r}

df_trees <- df_trees %>% 
  mutate(age = year(Sys.Date())-year)


hist(df_trees$age, main = 'Tree Age Distribution', xlab = 'Age', col = 'royalblue2') # using base r

```




Count the species
```{r}
df_trees %>%
  count(species_lat) %>%
  arrange(desc(n)) %>%
  top_n(n = 20)
```

Show number of distinct species in a street

```{r}
df_trees %>%
  group_by(address) %>%
  summarize(n = n_distinct(species_lat)) %>%
  arrange(desc(n))
```



# Analysis 1: evolution of planted species over time

To do this specific analysis we drop the trees, which have missing date data for obvious reasons. Based on our missing date data analysis in the data cleaning section, we are confident the data does not display any obvious signs of MNAR. 

Selected related columns

```{r}
df_trees_hasdate <- df_trees %>% 
  filter(!is.na(date)) %>% 
  select(c("tree_id", "species_lat", "year"))

str(df_trees_hasdate)
```
  
```{r}
df_trees_hasdate$species_lat <- as.factor(df_trees_hasdate$species_lat)
str(df_trees_hasdate)
```



To be able to decrease the number of species, we check the different species with a common general name (i.e Acacia, Magnolia)

```{r}
unique(df_trees_hasdate[grep(df_trees_hasdate$species_lat, pattern = "^Acacia", ignore.case = TRUE), 'species_lat'])
```


```{r}
unique(df_trees_hasdate[grep(df_trees_hasdate$species_lat, pattern = "^Magnolia", ignore.case = TRUE), 'species_lat'])
```

It seems that specific species can be grouped together with a general name. Since the general name is the first word of the species name, we will separate the species name into its words and then use the first word as the species general name. 


Separate the species_lat column into first and second part

```{r}
df_trees_hasdate <- df_trees_hasdate %>% 
  separate(species_lat, sep = " ", remove = FALSE, into = c('species_first', 'species_second', 'species_third'))
```

The code below shows that the number of unique species after recategorizing **species_lat** is 158.
```{r}
length(unique(df_trees_hasdate$species_first))
```


Create a count column to save number of planted species per year. 
```{r}
df_trees_hasdate_tidy <- df_trees_hasdate %>%
  group_by(year) %>%
  count(species_first) %>%
  arrange(year, desc(n)) 
```


Calculate the total number of planted trees per year and show it in a graph
```{r}
df_trees_hasdate_tidy %>%
  select(c('year', 'n')) %>%
  aggregate(by = list(df_trees_hasdate_tidy$year), FUN='sum') %>%
  ggplot(aes(x = Group.1, y = n)) + 
  geom_line() + 
  labs(title = 'Total Number of Planted Trees per Year', 
       x = 'Year', y = 'Total number of planted trees') +
  theme_minimal()
```


### Animated Graph

In this step, We’re going to filter our dataset to retain only the top 10 planted species for every given year. 

```{r}
df_trees_hasdate_formated <- df_trees_hasdate_tidy %>%
  group_by(year) %>%
  mutate(rank = rank(-n)) %>%
  filter(rank <=10) 
```


In this step, we will create individual static graphs for each year. 

```{r}
staticplot = ggplot(df_trees_hasdate_formated, aes(rank, group = species_first, 
                fill = as.factor(species_first), color = as.factor(species_first))) +
  geom_tile(aes(y = n/2,
                height = n,
                width = 0.9), alpha = 0.8, color = NA) +
  geom_text(aes(y = 0, label = paste(species_first, " ")), vjust = 0.2, hjust = 1) +
  geom_text(aes(y=n, label = n, hjust=0)) +
  coord_flip(clip = "off", expand = FALSE) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_reverse() +
  guides(color = FALSE, fill = FALSE) +
  theme(axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
         axis.title.y=element_blank(),
        legend.position="none",
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.grid.major.x = element_line( size=.1, color="grey" ),
        panel.grid.minor.x = element_line( size=.1, color="grey" ),
        plot.title=element_text(size=25, hjust=0.5, face="bold", colour="grey", vjust=-1),
        plot.subtitle=element_text(size=18, hjust=0.5, face="italic", color="grey"),
        plot.caption =element_text(size=8, hjust=0.5, face="italic", color="grey"),
        plot.background=element_blank(),
       plot.margin = margin(2,2, 2, 4, "cm"))
```


```{r}
anim = staticplot + transition_states(year, transition_length = 4, state_length = 1) +
  view_follow(fixed_x = TRUE)  +
  labs(title = 'Total Number of Planted Trees per Year: {closest_state}',  
       subtitle  =  "Top 10 Species")
```
  
For gif
```{r}
animate(anim, 200, fps = 3,  width = 1200, height = 1000, 
        renderer = gifski_renderer("gganim.gif"))

```

For Video
```{r}
# For MP4
for_mp4 <- animate(anim, 200, fps = 5,  width = 1200, height = 1000, renderer = av_renderer()) 
anim_save("animation.mp4", animation = for_mp4 )

```
  
## Alternative: cumulative total of top 5 species over time

find out which are the top 5 most planted species over entire time period. 'Trees' category will be omitted as it tells us nothing
about species

```{r}
df_agg <- aggregate(df_trees_hasdate_tidy$n, by=list(df_trees_hasdate_tidy$species_first), FUN='sum')
df_agg %>%
  arrange(desc(x)) %>%
  top_n(n=6)
```

create df to be used for cumulative summation
```{r}
df_top5 <- df_trees_hasdate_tidy %>%
  filter(species_first == 'Prunus' | species_first == 'Tristaniopsis' | species_first == 'Lophostemon' | species_first == 'Magnolia' | species_first =='Arbutus')
```


Calculate the cumulative summattion for each species. 
```{r}
df_top5_cumsum <- df_top5 %>% 
  group_by(species_first) %>% 
  mutate(csum = cumsum(n))
```

```{r}
#install.packages('transformr')
library('transformr')
```


Save the data type of **year** column as Date 
```{r}
df_top5_cumsum$year <- lubridate::ymd(df_top5_cumsum$year, truncated = 2L)
```


Create the static plot
```{r}
species_cumsum <- ggplot(df_top5_cumsum, aes(x=year,y=csum, group=species_first, col=factor(species_first))) + geom_line() + scale_color_manual(values = c('#e41a1c', '#377eb8', '#4daf4a', '#984ea3', '#ff7f00')) + theme_minimal() 
                                                                                      
```


Create the animated plot with transition 
```{r}
anim2 <- species_cumsum + transition_reveal(year) + view_follow(fixed_x = TRUE)  +
  labs(title = 'Cumulative Total of Top 5 Species over Time', x = 'Year', y = 'Cumulative Sum')
```

Animated graph (gif version): 

```{r}
animate(anim2, 200, fps = 3,  width = 1200, height = 1000, 
        renderer = gifski_renderer("gganim2.gif"))
```

  

#Analysis 2: 

Recommended Species (https://sfenvironment.org/sites/default/files/fliers/files/sf_tree_guide.pdf) : 


```{r}
recommended_sp <- c(' Japanese Blueberry Tree', ' Flaxleaf Paperbark', 
                            ' Red Flowering Gum', ' Flowering Cherry', ' Little Gem Magnolia')
```

<!-- to check the naming -->
```{r}
length(df_trees[grep(df_trees$species_nor, pattern = "Little Gem Magnolia", ignore.case = TRUE), 'species'])

```


```{r}
df_trees$species_nor <- as.factor(df_trees$species_nor)
```


```{r}
df_trees_filtered <- df_trees %>% 
  filter(species_nor %in% recommended_sp)
```


<!-- to check whether all recommended species are selected -->
```{r}
unique(df_trees_filtered$species_nor)
```


Get the coordinates of the whole dataset

```{r}
df_trees_filtered %>% 
  summarise(max_lon = max(longitude,  na.rm=TRUE), min_lon = min(longitude,  na.rm=TRUE),
            max_lat = max(latitude,  na.rm=TRUE), min_lat = min(latitude,  na.rm=TRUE))
```

```{r}
register_google('AIzaSyB2pTXikWRKM3XgN4WC-p-Fu2mOs8fZK88')
```

```{r}
san_francisco = c(left=-122.7, bottom = 37.7, right = -122.3, top = 37.85)
smap <- get_map(san_francisco, maptype = 'hybrid') %>% #get_stamenmap(san_francisco, maptype = 'watercolor')
  ggmap()
```

```{r}
smap + geom_point(data = df_trees_filtered, aes(x=longitude, y = latitude, col = species_nor)) + scale_color_manual(values = c('#e41a1c', '#377eb8', '#4daf4a', '#984ea3', '#ff7f00'))

```

Create a map with leaflet package
```{r}
#map center
mlong = -122.4446
mlat  = 37.75695

leaflet(data = df_trees_filtered, 
                width = '100%', 
                options = leafletOptions(minZoom = 9, maxZoom = 18)) %>% 
        addTiles() %>%
        setView(lng = mlong, lat = mlat, zoom = 11) %>%
        addCircleMarkers(lng = df_trees_filtered[df_trees_filtered$species_nor == ' Japanese Blueberry Tree', ]$longitude, 
                         lat = df_trees_filtered[df_trees_filtered$species_nor == ' Japanese Blueberry Tree', ]$latitude,
                         popup = df_trees_filtered[df_trees_filtered$species_nor == ' Japanese Blueberry Tree', ]$species_nor,
                         label = df_trees_filtered[df_trees_filtered$species_nor == ' Japanese Blueberry Tree', ]$species_nor, 
                         radius = 4)
```


Create a shiny app

```{r}
##doesnt work

#map center
mlong = -122.4446
mlat  = 37.75695

shinyApp(
  
  # Define the UI
  ui = fluidPage(
      
    # App title
    titlePanel("Trees of San Francisco"),
      
    # Sidebar layout with input and output definitions
    sidebarLayout(
        
      # Sidebar panel for inputs 
      sidebarPanel(
          
        # First input: Type of data
        selectInput(inputId = "species", #name of the input, widget
                      label = "Choose a species:",
                      choices = c('All Species', ' Japanese Blueberry Tree', ' Flaxleaf Paperbark', 
                                  ' Red Flowering Gum', ' Flowering Cherry', ' Little Gem Magnolia'), 
                      selected = ' Japanese Blueberry Tree'
        )),
        
      # Main panel for displaying outputs
      mainPanel(
          
        # Hide errors
        tags$style(type = "text/css",
                     ".shiny-output-error { visibility: hidden; }",
                     ".shiny-output-error:before { visibility: hidden; }"),
          
        # Output: interactive world map
        plotOutput("map")
          
        )
      )
    ),
    
  # Define the server
  server = function(input, output) {
    
    filteredData <- reactive({
      if (input$species == "All Species") {
        df_trees_filtered
      } else {
        filter(df_trees_filtered, species_nor == input$species)
      }
    })
      
    output$map <- renderLeaflet({
          
        leaflet(filteredData(), 
                width = '100%', 
                options = leafletOptions(minZoom = 9, maxZoom = 18)) %>% 
        addTiles() %>%
        setView(lng = mlong, lat = mlat, zoom = 11) %>%
        addCircleMarkers(lng = filteredData()$longitude, 
                           lat = filteredData()$latitude,
                           popup = filteredData()$species_nor,
                           #label = ~species_nor, 
                           radius = 4)
    })
  }
)
  


```




```{r}
runExample("03_reactivity")
```



- map shows the top 5(/10) most planted species on a map


#Analysis 3:
- another map with rare (n=1/below 5) species 











