---
title: "Trees of San Francisco"
author:
- Gökce Ergün (19-865-526)
- Nicolas Garzon
date: "2/3/2021"
output:
  html_document: default
  pdf_document: default
  word_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, message = FALSE)
```

**The client**: The city of San Francisco 

**The scenario**: The city of SF is looking to make use of a data set dating back some 50 years that has never been put to any use
and where they have lost the overview. We are a data consultants who were given the data set to create a preliminary exploratory analyis type report as a trail with view to being hired on a longer term project with the [data set](https://github.com/rfordatascience/tidytuesday/tree/master/data/2020/2020-01-28).


Libraries that are used
```{r}
#install.packages('dplyr')
library(dplyr)
#install.packages('lubridate')
library(lubridate)
#install.packages('tidyr')
library(tidyr)
#install.packages('ggplot2')
library(ggplot2)
library(tidyverse)
#install.packages('gganimate')
library(gganimate)
#install.packages('gifski')
library(gifski)
#install.packages('av')
library(av)
#install.packages('ggmap')
library(ggmap)
#install.packages('shiny')
library(shiny)
#install.packages('leaflet')
library(leaflet)
```



# Reading the data
```{r}
df_trees_orig <- read.csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-28/sf_trees.csv')

```

We create a copy of data to preserve its original form in case we want to look at it for reference. 
```{r}
df_trees <- df_trees_orig
```

# Data Cleaning 

We look at the data structure, and see all the string variables are saved as char. We would rather have them as factors for potential modelling use.
```{r}
str(df_trees)
```

Changing columns with type character to type factor. 

NOTE: You need double brackets because single bracket indexing returns a df object,
which always evaluates to FALSE. Double brackets returns you the column as a vector whose data type can be tested. 
Alternatively, we can use single brackets but explicitly calling all rows and columns i.e [,i]
```{r}

for (i in 1:12){ 
    if (is.character(df_trees[,i]) == TRUE) {
    df_trees[,i]<-as.factor(df_trees[,i])
  }
  }

```

<!--check the changes worked-->

```{r, include=FALSE}
str(df_trees)
``` 

**Date** column has character data type. We change it to date. 
```{r}
# df_trees$date <- as.Date(df_trees$date,format="%Y-%m-%d") #using base r

df_trees$date <- lubridate::ymd(df_trees$date) #using lubridate
```

Save the planted year of the tree as a separate column
```{r}
# df_trees$year <- format(df_trees$date, "%Y")  #using base r

df_trees$year <- year(df_trees$date) # using lubridate
class(df_trees$year)
```

Take a look at the format of species name in species column. 
```{r}
head(unique(df_trees$species))

```

we see that for each **species**, the species column reports both the latin and common names. For future analysis, it might be more 
convenient to split the latin and common names into respective columns. 

Separate the species column into latin and common name columns.
```{r}
df_trees <- df_trees %>% 
  separate(species, sep = "::", remove = FALSE, into = c('species_lat', 'species_nor'))
```

<!--check the name split worked-->
```{r, echo=FALSE, results='hide'}
head(df_trees[,c('species','species_lat','species_nor')])
```

Check the percentage of missing values in each column

```{r}
for (col in colnames(df_trees)){
  print(paste(col, "=", mean(is.na(df_trees[[col]])))) #mean takes percentage here because adds up all TRUES (since their value is 1)
                                                       #and divides by total number of values (TRUE + FALSE)
}
```

Column with most significant amount of missing data is the **date** column.

We see almost every species has missing values (517 out of 571 species) so there is at least no clear indication that the data is missing not at random.
```{r}
df_trees %>% filter(is.na(date))%>%
  select(species)%>%
  unique() %>% dim()
 
```

Further analysis of missing date values: percentage of observations with missing date values, per species.
```{r}
df_trees_datena <- df_trees %>% filter(is.na(date)) %>% 
 count(species)
 
df_trees_species_count <- df_trees %>% count(species)

df_trees_merge <- dplyr::left_join(df_trees_species_count, df_trees_datena, by="species", suffix = c("total", "date.na"))
df_trees_merge$percent.missing <- df_trees_merge$ndate.na / df_trees_merge$ntotal
df_trees_merge %>%
  arrange(desc(ntotal), desc(percent.missing)) %>%
  head()

```


We look at the total number of observations we would lose if we drop species where missing 100% of date information. And the number is very small relative to total number of observations (n = 192987) 

```{r}
df_trees_merge %>% filter(percent.missing==1) %>% select(ntotal) %>%
  sum()
```

How the percentage of missing date values per species is distributed. All species that have 100% of date data missing have a very low number of observations (n <=100).    

```{r}
df_trees_merge %>% 
  ggplot(aes(percent.missing)) + geom_histogram(fill='skyblue4') + 
  labs(title='Percentage of Missing Date Values per Species') +
  xlab('Percentage of missing date') + ylab('Count') + 
  scale_x_continuous(breaks=seq(0,1, .10)) +theme_minimal()

```


Therefore, we now filter the cases based on this 100 ntotal threshold to get a cleaner histogram. 

```{r}
df_trees_merge %>% filter(ntotal>100) %>%
  ggplot(aes(percent.missing))+geom_histogram(fill='skyblue4') + 
  labs(title='Percentage of Missing Date Values per Species') +
  xlab('Percentage of missing date') + ylab('Count') + 
  scale_x_continuous(breaks=seq(0,1, .10)) + theme_minimal()

```


# Exploratory Analysis

Checking year of plantation range. most of the trees (50%) were planted between 2001 and 2020. However, it is likely that many of
the older trees simply were not recorded so they either dont appear in our data or have a missing date value. 
```{r}
df_trees %>%
  select(year)%>%
  quantile(na.rm = TRUE)
  
```

Create a new column that shows age using the year column to calculate it. Use the more modern approach mutate(). Then use age column 
to get an overview of the distribution of tree ages via a histogram.

```{r}

df_trees <- df_trees %>% 
  mutate(age = year(Sys.Date())-year)


hist(df_trees$age, main = 'Tree Age Distribution', xlab = 'Age', col = 'skyblue4') # using base r

```


Count the species
```{r}
df_trees %>%
  count(species_lat) %>%
  arrange(desc(n)) %>%
  top_n(n = 20)
```

Show number of distinct species in a street

```{r}
df_trees %>%
  filter(address != 'NA') %>%
  group_by(address) %>%
  summarise(n = n_distinct(species_lat)) %>%
  arrange(desc(n)) %>%
  top_n(n = 10)
```



# Analysis 1: Evolution of planted species over time

To do this specific analysis we drop the trees, which have missing date data for obvious reasons. Based on our missing date data analysis in the data cleaning section, we are confident the data does not display any obvious signs of MNAR. 

Selected related columns

```{r}
df_trees_hasdate <- df_trees %>% 
  filter(!is.na(date)) %>% 
  select(c("tree_id", "species_lat", "year"))

str(df_trees_hasdate)
```
 
**Species_lat** is saved as a factor.   
```{r}
df_trees_hasdate$species_lat <- as.factor(df_trees_hasdate$species_lat)
```


To make the number of species more manageable, we group them by family names (i.e Acacia, Magnolia)

```{r}
unique(df_trees_hasdate[grep(df_trees_hasdate$species_lat, pattern = "^Acacia", ignore.case = TRUE), 'species_lat'])
```


```{r}
unique(df_trees_hasdate[grep(df_trees_hasdate$species_lat, pattern = "^Magnolia", ignore.case = TRUE), 'species_lat'])
```

It seems that specific species can be grouped together with a general name. Since the general name is the first word of the species name, we will separate the species name into its words and then use the first word as the species general name. 


Separate the species_lat column into its parts

```{r}
df_trees_hasdate <- df_trees_hasdate %>% 
  separate(species_lat, sep = " ", remove = FALSE, into = c('species_first', 'species_second', 'species_third'))
```

The code below shows that the number of unique species after recategorizing **species_lat** is 158.
```{r}
length(unique(df_trees_hasdate$species_first))
```


Create a count column to save number of planted trees by species, per year. 
```{r}
df_trees_hasdate_tidy <- df_trees_hasdate %>%
  group_by(year) %>%
  count(species_first) %>%
  arrange(year, desc(n)) 
```


As a bit of side note we look at the total number of planted trees per year regardless of species to get a sense of overall tree planting
```{r}
df_trees_hasdate_tidy %>%
  select(c('year', 'n')) %>%
  aggregate(by = list(df_trees_hasdate_tidy$year), FUN='sum') %>%
  ggplot(aes(x = Group.1, y = n)) + 
  geom_line() + 
  labs(title = 'Total Number of Planted Trees per Year', 
       x = 'Year', y = 'Total number of planted trees') +
  theme_minimal()
```


### Animated Graph

In this step, We’re going to filter our data set to retain only the top 10 planted species for every given year to keep the visuals
understandable. Also we drop the species labelled Tree(s) as this obviously tells us nothing about the species.

```{r}
df_trees_hasdate_formated <- df_trees_hasdate_tidy %>% subset(species_first != 'Tree(s)') %>% 
  group_by(year) %>%
  mutate(rank = rank(-n)) %>%
  filter(rank <=10) 
```


In this step, we will create individual static graphs for each year. 

```{r cache = TRUE}
staticplot = ggplot(df_trees_hasdate_formated, aes(rank, group = species_first, 
                fill = as.factor(species_first), color = as.factor(species_first))) +
  geom_tile(aes(y = n/2,
                height = n,
                width = 0.9), alpha = 0.8, color = NA) +
  geom_text(aes(y = 0, label = paste(species_first, " ")), vjust = 0.2, hjust = 1) +
  geom_text(aes(y=n, label = n, hjust=0)) +
  coord_flip(clip = "off", expand = FALSE) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_reverse() +
  guides(color = FALSE, fill = FALSE) +
  theme(axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
         axis.title.y=element_blank(),
        legend.position="none",
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.grid.major.x = element_line( size=.1, color="grey" ),
        panel.grid.minor.x = element_line( size=.1, color="grey" ),
        plot.title=element_text(size=25, hjust=0.5, face="bold", colour="grey", vjust=-1),
        plot.subtitle=element_text(size=18, hjust=0.5, face="italic", color="grey"),
        plot.caption =element_text(size=8, hjust=0.5, face="italic", color="grey"),
        plot.background=element_blank(),
       plot.margin = margin(2,2, 2, 4, "cm"))
```


```{r cache=TRUE}
anim = staticplot + transition_states(year, transition_length = 4, state_length = 1) +
  view_follow(fixed_x = TRUE)  +
  labs(title = 'Total Number of Planted Trees per Year: {closest_state}',  
       subtitle  =  "Top 10 Species")
```
  
For gif
```{r cache = TRUE}
animate(anim, 200, fps = 2,  width = 1200, height = 1000, 
        renderer = gifski_renderer("gganim.gif"))

```

### Alternative: Cumulative total of top 5 species over time

find out which are the top 5 most planted species over entire time period. 'Trees' category will be omitted as it tells us nothing
about species

```{r}
df_agg <- aggregate(df_trees_hasdate_tidy$n, by=list(df_trees_hasdate_tidy$species_first), FUN='sum')
df_agg %>%
  arrange(desc(x)) %>%
  top_n(n=6)
```

create df to be used for cumulative summation
```{r}
df_top5 <- df_trees_hasdate_tidy %>%
  filter(species_first == 'Prunus' | species_first == 'Tristaniopsis' | species_first == 'Lophostemon' | species_first == 'Magnolia' | species_first =='Arbutus')
```


Calculate the cumulative summation for each species. 
```{r}
df_top5_cumsum <- df_top5 %>% 
  group_by(species_first) %>% 
  mutate(csum = cumsum(n))
```

Save the data type of **year** column as Date 
```{r}
df_top5_cumsum$year <- lubridate::ymd(df_top5_cumsum$year, truncated = 2L)
```

Create the static plot
```{r cache=TRUE}
species_cumsum <- ggplot(df_top5_cumsum, aes(x=year,y=csum, group=species_first, col=factor(species_first))) + geom_line() + scale_color_manual(values = c('#e41a1c', '#377eb8', '#4daf4a', '#984ea3', '#ff7f00')) +  
  theme_minimal(base_size = 22) 
                                                                                      
```


Create the animated plot with transition 
```{r cache=TRUE}
anim2 <- species_cumsum + transition_reveal(year) + view_follow(fixed_x = TRUE)  +
  labs(title = 'Cumulative Total of Top 5 Species over Time', x = 'Year', y = 'Cumulative Sum', colour = 'Top 5 Species') 
```
 
Animated graph (gif version): 

```{r cache=TRUE}
animate(anim2, 200, fps = 5,  width = 1200, height = 1000, 
        renderer = gifski_renderer("gganim2.gif"))
```

  

### Analysis 2: 

map of recommended species per  [Source](https://sfenvironment.org/sites/default/files/fliers/files/sf_tree_guide.pdf): 


```{r}
recommended_sp <- c(' Japanese Blueberry Tree', ' Flaxleaf Paperbark', ' Red Flowering Gum', ' Flowering Cherry', 
                    ' Little Gem Magnolia', ' Southern Magnolia', ' Weeping Bottlebrush', ' Hybrid Strawberry Tree', 
                    ' Primrose Tree', ' Brisbane Box', ' Bronze Loquat', ' Peppermint Willow', ' Mediterranean Fan Palm', 
                    ' Fruitless Olive', ' Chilean Soapbark', " Small-leaf Tristania 'Elegant'", ' Chinese Pistache', 
                    ' Trident Maple', ' Chinese Elm', ' Cork Oak', ' Ginkgo: Autumn Gold', ' Fairmont Ginkgo', 
                    ' Ginkgo: Saratoga', ' Autumn Sentinel Ginkgo'
                    )
```

<!-- to check the naming -->
```{r echo=FALSE,results='hide'}
unique(df_trees[grep(df_trees$species_nor, pattern = "Peppermint Willow", ignore.case = TRUE), 'species'])

```

**Species_nor** is saved as a factor. 
```{r}
df_trees$species_nor <- as.factor(df_trees$species_nor)
```

Recommended species are filtered.
```{r}
df_trees_recommended <- df_trees %>% 
  filter(species_nor %in% recommended_sp) %>%
  select(c("species_nor", 'latitude', "longitude", 'dbh', 'age'))
```


<!-- to check whether all recommended species are selected -->
```{r}
unique(df_trees_recommended$species_nor)
```


Get the coordinates of the whole dataset

```{r}
df_trees_recommended %>% 
  summarise(max_lon = max(longitude,  na.rm=TRUE), min_lon = min(longitude,  na.rm=TRUE),
            max_lat = max(latitude,  na.rm=TRUE), min_lat = min(latitude,  na.rm=TRUE))
```

Create a map with leaflet() package
```{r fig.height=12, fig.width=8}
#map center
mlong = -122.4446
mlat  = 37.72

pal <- colorFactor('Paired', domain = recommended_sp)

leaflet(data = df_trees_recommended, 
                width = '100%', 
                height = 800,
                options = leafletOptions(minZoom = 9, maxZoom = 18)) %>% 
        addTiles() %>%
        setView(lng = mlong, lat = mlat, zoom = 12) %>%
        addCircleMarkers(lng = df_trees_recommended$longitude, 
                         lat = df_trees_recommended$latitude,
                         color = ~pal(df_trees_recommended$species_nor),
                         popup = df_trees_recommended$species_nor,
                         label = df_trees_recommended$species_nor, 
                         radius = 4) %>%
      addLayersControl(overlayGroups = df_trees_recommended$species_nor) ## add layers control
```

But map is too chaotic given large number of species...sooo Shiny to the rescue:

Create a shiny app version of the map above, with a check-box to look at each species at a time.

<<link>>

## Model ##

Categorize the species based on their heights

```{r}
small_trees <- c(' Trident Maple', ' Weeping Bottlebrush', ' Mediterranean Fan Palm', ' Bronze Loquat', ' Little Gem Magnolia',
                 ' Flowering Cherry')

medium_trees <- c(' Hybrid Strawberry Tree', " Small-leaf Tristania 'Elegant'", ' Peppermint Willow', ' Japanese Blueberry Tree', 
                 ' Flaxleaf Paperbark', ' Fruitless Olive', ' Chinese Pistache', ' Red Flowering Gum', ' Primrose Tree')
                    
large_trees <- c(' Brisbane Box', ' Southern Magnolia', ' Cork Oak', ' Chilean Soapbark', ' Ginkgo: Autumn Gold', 
                 ' Fairmont Ginkgo', ' Ginkgo: Saratoga', ' Autumn Sentinel Ginkgo', ' Chinese Elm') 

```

Create a **height** column. 

```{r}
df_trees_recommended <- df_trees_recommended %>%
  mutate(height = case_when(species_nor %in% small_trees ~ 'small', 
                          species_nor %in% medium_trees ~ 'medium',
                          species_nor %in% large_trees ~ 'large'))
```

Create a graph

```{r}
df_trees_recommended %>%
  filter(age != 'NA') %>% 
  ggplot(aes(x = age, y = dbh)) + labs(x = 'Age', y = 'Diameter at breast height') +
  geom_point() + facet_wrap(.~height) + geom_smooth() + theme_minimal()

```

Predict the diameter of tree at breast height (**dbh**)

```{r}
library(mgcv)
gam.1 <- gam(dbh ~ s(age) + height, data = df_trees_recommended)
summary(gam.1)
```



# Analysis 3:
- another map with rare (n=1) species

create list of rare species (n = 1) with name matching original df_trees 
```{r}
species_rare <-  df_trees_species_count%>%
  filter(n == 1) %>%
  separate(species, sep = "::", remove = FALSE, into = c('species_lat', 'species_nor'))

```

Use list of rare species to subset original df_trees and also group them by common first name (i.e family name) to reduce species number
```{r}
df_trees_rare <- df_trees %>%
  filter(species_nor %in% species_rare$species_nor) %>%
  separate(species_lat, sep = " ", remove = FALSE, into = c('species_first', 'species_second', 'species_third'))

  
```


Plot map using leaflet, to create a hover-over dot pop up with species name, given not enough colors to differentiate so many species
```{r}
#map center
mlong = -122.4446
mlat  = 37.72


rare_species_vec <- unique(df_trees_rare$species_lat)
pal <- colorFactor('Paired', domain = rare_species_vec)

leaflet(data = df_trees_rare, 
                width = '100%', height = 800,
                options = leafletOptions(minZoom = 9, maxZoom = 18)) %>% 
        addTiles() %>%
        setView(lng = mlong, lat = mlat, zoom = 12) %>%
        addMarkers(lng = df_trees_rare$longitude, 
                         lat = df_trees_rare$latitude,
                         icon = list(
                            iconUrl = 'https://icons.iconarchive.com/icons/matthew-kelleigh/mac-town-vol2/32/Tree-1-icon.png',
                              #'https://icons.iconarchive.com/icons/pixelkit/flat-jewels/512/Tree-icon.png',
                            iconSize = c(30, 30)
                         ), 
                         popup = df_trees_rare$species_lat,
                         label = df_trees_rare$species_lat)
                         #radius = 4,
                         #fillOpacity = .5)

```
